Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-07-25 chap-intro.tex
Move contributions into parts
Already done for Part I earlier, do it for Part II.

\pg{Rewrite}
This thesis is divided into two parts:
\begin{itemize}
\item In \cref{part:ch-aosd13}, we optimize collection
  queries by static program transformation, based on a set of rewrite
  rules~\citep{GiarrussoAOSD13}.
\item In \cref{part:incr}, we incrementalize programs by transforming them to
  new programs. This thesis presents the first approach that
  handles higher-order programs through program transformation; hence, while
  our main examples use collection queries, we phrase the work in terms of
  $\lambda$-calculi with unspecified primitives~\citep*{CaiEtAl2014ILC}.
  In \cref{ch:cts}, we extend ILC with a further program transformation
  step, so that base programs can store intermediate results and derivatives can
  reuse them, but without resorting to dynamic memoization and necessarily
  needing to look results up at runtime. To this end, we build on work by
  \citet{Liu00} and extend it to a higher-order, typed setting.
\end{itemize}

@@

%\Cref{part:ch-aosd13} makes the following contributions:
\pg{TODO:contributions of other parts.}
\pg{Contributions: maybe in first chapters?}

\Cref{part:incr} makes the following contributions:
\begin{itemize}
\item We present a novel mathematical theory of changes and derivatives, which
  is more general than other work in the field because changes are first-class
  entities, they are distinct from base values and they are defined also for
  functions.
  %(\cref{sec:1st-order-changes}).
  %KO: I think the next sentence cannot be understood at this point.
  %We introduce changes for complex types, defined compositionally.
%
\item We present the first approach to incremental computation for pure
  $\lambda$-calculi by a source-to-source transformation, $\DERIVE$, that
  requires no run-time support. The transformation produces an incremental
  program in the same language; all optimization techniques for the original
  program are applicable to the incremental program as well.

  % \pg{ Maybe comment in? Not sure, which might be telling.}
  % Since our incremental programs use no impure features, they are
  % especially amenable to further optimizations, making this approach
  % very suitable for further research.

%KO: commented this out. I think the purity is not important enough
%to deserve another sentence here, since we only vaguely hint
%at "further research".
% KO: Let's have one bullet point per section. Also, a conjecture
% sounds like a rather weak contribution
%\item We argue that incrementalization is efficient on
%  \emph{self-maintainable programs}, and discuss how further research on
%  static or dynamic memoization can speed up a larger class of programs (\cref{sec:performance-cons}).
%  \pg{This contribution references text which is now commented
%    out. I believe the text should be brought back in.}
%
\item We prove that our incrementalizing transformation $\DERIVE$
is correct by a novel machine-checked logical relation
proof, mechanized in Agda.

\item While we focus mainly on the theory of changes
and derivatives, we also perform a performance case study.
We implement the derivation transformation in Scala,
with a plug-in architecture that can be extended with new base
types and primitives. We define a plugin with support for
different collection types and use the plugin to
incrementalize a variant of the MapReduce programming model~\citep{Lammel07}.
  Benchmarks show that on this program,
  incrementalization can reduce asymptotic complexity and can turn $O(n)$
  performance into $O(1)$, improving running time by over 4
  orders of magnitude on realistic inputs (\cref{sec:applying}).
\end{itemize}

\pg{Describe \cref{ch:cts}}
\pg{Paste contributions from paper}
\pg{Maybe defunc chapter?}

\pg{WIP}
Other chapters propose further contributions.
\Cref{ch:bsos} proposes the first correctness proofs for ILC via operational
methods and (step-indexed) logical relations, for simply-typed $\lambda$-calculus
(without and with general recursion) and for untyped $\lambda$-calculus.

\Cref{ch:diff-parametricity-system-f} outlines how to extend differentiation to
System F and suggests a proof avenue. Differentiation for System F requires a
generalization of change structures to changes across different types
(\cref{sec:param-derive-changes-across-types}) that appears of independent
interest, though further research remains to be done.

@@

\subsection{Navigating this thesis}
\pg{WIP!!}
Many chapters are independent from each other, with a few exceptions.
\Cref{part:ch-aosd13} can be read independently from the rest.
% Write this even though maybe I restructure the parts.
\Cref{part:incr} describes incrementalization.

\paragraph{Differentiation}
\Crefrange{sec:intro}{sec:term-reasoning} form the core of incrementalization
theory, with other chapters building on top of them.

\Cref{sec:intro} and \cref{ch:diff-examples} introduce the overall approach to
incrementalization informally.

\Cref{ch:derive-formally,ch:change-theory} show that incrementalization using
ILC is correct. Building on a set-theoretic semantics, these
\lcnamecrefs{ch:derive-formally} also develop the theory
underlying this correctness proofs and further results.

Equational reasoning on terms is then developed
in \cref{sec:term-reasoning}. These chapters contain full formal proofs: readers
are welcome to skip or skim those proofs where appropriate. We strive not to
hide important definitions inside theorems, and to give stand-alone definitions
and theorem statements.

Later chapters are again independent from each other.

\paragraph{Performance of differentiated programs}
\Cref{sec:applying} studies how to apply differentiation (as introduced in
previous chapters) to incrementalize a case study and empirically evaluates
performance speedups.

\paragraph{Differentiation with static caching}
\pg{Explain label}
\Cref{ch:cts} is mostly self-contained, even though it builds on the rest of the
material; it summarizes the basics of ILC in \cref{sec:ilc-background}.
Terminology in that chapter is sometimes subtly different from the rest of the thesis.

\pg{At least in more formal chapters, all definitions, notations, and so on are
  clearly highlighted and presented as standalone definitions for ease of reference.}

\pg{Missing chapters}

>>>>

This thesis is divided into two parts:
\begin{itemize}
\item In \cref{part:ch-aosd13}, we optimize collection
  queries by static program transformation, based on a set of rewrite
  rules~\citep{GiarrussoAOSD13}.
\item In \cref{part:incr}, we incrementalize programs by transforming them to
  new programs. This thesis presents the first approach that
  handles higher-order programs through program transformation; hence, while
  our main examples use collection queries, we phrase the work in terms of
  $\lambda$-calculi with unspecified primitives~\citep*{CaiEtAl2014ILC}.
  In \cref{ch:cts}, we extend ILC with a further program transformation
  step, so that base programs can store intermediate results and derivatives can
  reuse them, but without resorting to dynamic memoization and necessarily
  needing to look results up at runtime. To this end, we build on work by
  \citet{Liu00} and extend it to a higher-order, typed setting.
\end{itemize}

@@

Contributions of \cref{part:ch-aosd13} are listed in \cref{sec:aosd13-contributions}.
Contributions of \cref{part:incr} are listed at the end of
\cref{sec:incr-contributions}.
 %\Cref{part:ch-aosd13} makes the following contributions:
%\pg{Contributions: maybe in first chapters?} % Yes

@@

\subsection{Navigating this thesis}
The two parts of this thesis, while related by the common topic of collection
queries, can be read mostly independently from each other.
Summaries of the two parts are given respectively in
\cref{sec:navigating-aosd13,sec:navigating-incr}.

% Many chapters are independent from each other, with a few exceptions.
% \Cref{part:ch-aosd13} can be read independently from the rest.
% % Write this even though maybe I restructure the parts.
% \Cref{part:incr} describes incrementalization.
